Characterizing Stellar Streams by ML

### 2023-08-21 src/tex/ms.tex
[showyourwork] overleaf sync

        The paper is organized as follows.
        In \autoref{sec:method} we introduce our method for characterizing stream tracks and computing stellar membership.
        We discuss the data in \autoref{sub:framing_the_data} and build a probabilistic framework to model the stream and background in \autoref{sub:likelihood_setup}-\autoref{sub:priors}.
        In \autoref{sec:results} we apply our method,
        % TODO: rephrase the mock data
        first to simulated data (\autoref{sub:mock_data}) to demonstrate the model with a known ground truth dataset,
        then to the stellar streams \stream{GD-1} (\autoref{sec:results_gd1}) and \stream{Pal\,5} (\autoref{sub:pal5}).
        We compare our approach to characterizing stellar streams to other methods in \autoref{sec:comparison}.
        We conclude in \autoref{sec:conclusions}, discussing the results of this work and future directions.

@@

        \begin{figure}
            \centering
            \includegraphics[scale=0.5]{static/model/DAG_StreamDensity.pdf}
            \caption{Probabilistic Graphical Model of the Mixture Density Network (MDN)}
            \todo[inline]{todo}
            \begin{enumerate}
                \item remake this example this with \url{https://github.com/daft-dev/daft}.
            \end{enumerate}
            \label{fig:DAG}
        \end{figure}

@@

            The photometric probability, \autoref{eq:photometric_probability} includes the term $\prior(\gamma, \nth{\mbs{m}}, \phi_1)$, which encodes all priors on both the isochrone as well as observational constraints. It is safe to assume that the properties of the isochrone are conditionally independent from observational constraints; namely
            \begin{equation} \label{eq:isochrone_prior_components}
                \prior(\gamma, \nth{\mbs{m}}, \phi_1) = \prior_{\cal{I}}(\gamma, \phi_1) \prior_{obs}(\nth{\mbs{m}}, \phi_1)
            \end{equation}
            where $\prior_{\cal{I}}$ and $\prior_{obs}$ are the isochrone and observational priors, respectively.
            \autoref{eq:photometric_probability} already encodes the photometric distribution -- the ``track" -- of the isochrone model at each point along the track, but not the amplitudes, the relative probability of points. Physically, this amplitude is the the present-day mass function (PDMF). We include the PDMF as $\prior_{\cal{I}}(\gamma, \phi_1)$. $\prior_{\cal{I}}$ is a joint distribution, depending on $\gamma$ since it is the $\gamma$-dependent distribution amplitude, and also on $\phi_1$. This latter dependence is confusing termed the ``mass function of the stream", which is the expected variation in the average stellar mass along the stream, due to e.g. mass segregation in the progenitor \TODO{[cite]}. To date this effect has not been observed \TODO{[cite]}, so in practice we drop the $\phi_1$ dependence -- $\prior_{\cal{I}}(\gamma, \phi_1) = \prior_{\cal{I}}(\gamma)$ -- assuming that the PDMF is constant along the stream.

            % link to the code!
            In the code we offer a few ansatz for the PDMF, though any user-defined function may be used instead. The default assumption, which is known to be non-physical is a uniform distribution over $\gamma$. Similarly are truncated uniform distributions, considering only a segment of the isochrone, e.g. the main sequence turnoff (MSTO). We include also a Kroupa initial mass function (IMF) \citep{2001MNRAS.322..231K}. The Kroupa IMF, and all physically-motivated IMFs,
            have an much larger fraction of low-mass stars then high-mass stars, the latter being less likely to form. This is observed in streams, particularly their progenitors. However the Kroupa IMF, and all IMF, are poor fits to the observed PDMF as stream progenitors like globular clusters are extremely old ($\sim10$ Gyr) while an IMF is the zero-age distribution.
            How then should the PDMF $\prior_{\cal{I}}$ be treated?
            % ⬆ rephrase previous sentence
            One approach to the PDMF is to use a more informed distribution, for example time-evolving an IMF, either analytically or by simulation.
            Using any fixed distribution is simple in implementation, however the choice of distribution is not \TODO{[cite]}.
            There is large uncertainty in the PDMF of most globular clusters.
            % ⬆ rephrase previous sentence

            Similarly, the PDMF might be incorporated into the model,
            parameterized by some distribution and fit simultaneously to the distance modulus and other $\cmp{m}{\mbs{\theta}}$.
            Using a fixed distribution is a simple matter, however 
            \TODO{[cite]}

@@

    In this section we present an application of our model to a synthetic stream with known ground truths, and demonstrate the model's ability to characterize stream density variations and membership probabilities in an unsupervised manner. We then apply the method to real stellar streams in \autoref{sec:results_gd1} and \autoref{sub:pal5}.

@@

        We now apply our density modeling approach to the stellar stream GD-1. We discuss the data selection in \autoref{ssec:results_gd1_data} and results in \autoref{ssub:trained_gd1}.

        \subsubsection{Data Selection} \label{ssec:results_gd1_data}

@@

        \subsubsection{Model Specification}\label{sec:GD_1_ModelSpecification}

@@

        \subsubsection{Trained Density Model}\label{ssub:trained_gd1}

@@

    \subsection{Palomar 5} \label{sub:pal5}

@@

    % subsection pal5 (end)

>>>>

        The paper is organized as follows.
        In \autoref{sec:method} we introduce our method for characterizing stream tracks and computing stellar membership.
        We discuss the data in \autoref{sub:framing_the_data} and build a probabilistic framework to model the stream and background in \autoref{sub:likelihood_setup}-\autoref{sub:priors}.
        In \autoref{sec:results_mock} we apply our method to simulated data (\autoref{sub:mock_data}) to demonstrate the model with a known ground truth dataset,
        then to the stellar streams \stream{GD-1} (\autoref{sec:results_gd1}) and \stream{Pal\,5} (\autoref{sec:results_pal5}).
        We compare our approach to characterizing stellar streams to other methods in \autoref{sec:comparison}.
        We conclude in \autoref{sec:conclusions}, discussing the results of this work and future directions.

@@

        \begin{figure}
            \centering
            \includegraphics[scale=0.5]{static/model/DAG_StreamDensity.pdf}
            \caption{Probabilistic Graphical Model of the Mixture Density Network (MDN)}
            \todo[inline]{TODO:}
            \begin{enumerate}
                \item remake this example this with \url{https://github.com/daft-dev/daft}. See \url{https://en.wikipedia.org/wiki/Mixture_model} for examples.
            \end{enumerate}
            \label{fig:DAG}
        \end{figure}

@@

            The photometric probability, \autoref{eq:photometric_probability}, includes the term $\prior(\gamma, \nth{\mbs{m}}, \phi_1)$, which encodes all priors on both the isochrone as well as observational constraints. We assume that the properties of the isochrone are conditionally independent from observational constraints; namely
            \begin{equation} \label{eq:isochrone_prior_components}
                \prior(\gamma, \nth{\mbs{m}}, \phi_1) = \prior_{\cal{I}}(\gamma, \phi_1) \prior_{obs}(\nth{\mbs{m}}, \phi_1)
            \end{equation}
            where $\prior_{\cal{I}}$ and $\prior_{obs}$ are the isochrone and observational priors, respectively.
            \autoref{eq:photometric_probability} already encodes the photometric distribution -- the ``track" -- of the isochrone model at each point along the track, but not the amplitudes, the relative probability of points. Physically, this amplitude is the the present-day mass function (PDMF). We The PDMF $\prior_{\cal{I}}(\gamma, \phi_1)$ is a joint distribution, depending on $\gamma$ since it is the $\gamma$-dependent distribution amplitude, and also on $\phi_1$. This latter dependence is confusingly termed the ``mass function of the stream", which is the expected variation in the stellar mass distrubion along the stream, arising due to e.g. mass segregation in the progenitor \TODO{[cite]}. To date this variation along the stream has not been observed \TODO{[cite]}, so in practice we drop the $\phi_1$ dependence -- $\prior_{\cal{I}}(\gamma, \phi_1) = \prior_{\cal{I}}(\gamma)$ -- assuming that the PDMF is constant along the stream.

            % link to the code!
            In the code we offer a few ansatz for the PDMF, though any user-defined function may be used instead. The default assumption, known to be non-physical, is a uniform distribution over $\gamma$. Similarly non-physical are truncated uniform distributions, considering only a segment of the isochrone, e.g. the main sequence turnoff (MSTO). We include also a Kroupa initial mass function (IMF) \citep{2001MNRAS.322..231K}. The Kroupa IMF, and all physically-motivated IMFs,
            have an much larger fraction of low-mass stars then high-mass stars, the latter being less likely to form. This is observed in streams, particularly their progenitors. However the Kroupa IMF, and all IMF, are poor fits to the observed PDMF as stream progenitors like globular clusters are extremely old ($\sim10$ Gyr) while an IMF is the zero-age distribution.
            How then should the PDMF $\prior_{\cal{I}}$ be treated?
            % ⬆ rephrase previous sentence
            One approach to the PDMF is to use a more informed distribution, for example time-evolving an IMF, either analytically or by simulation, to the age of the stream's stellar population.
            As discussed previously, globular clusters are known to be poorly fit by theoretic isochrones with realistic inputs (e.g. age, $Z$). Thus fitting an isochrone's track to the data means its age, metallicity and therefore PDMF
            are not well fit \TODO{[cite]}. Moreover, for many streams, e.g. those lacking known progenitors, the PDMF is observationally poorly constrained.
            In short, using any fixed distribution is simple in implementation, however the choice of distribution is not.

            Rather than using a fixed \textit{a priori} 
            Similarly, the PDMF might be incorporated into the model,
            parameterized by some distribution and fit simultaneously to the distance modulus and other $\cmp{m}{\mbs{\theta}}$.
            Using a fixed distribution is a simple matter, however 
            \TODO{[cite]}

@@

    In this section we present an application of our model to a synthetic stream with known ground truths, and demonstrate the model's ability to characterize stream density variations and membership probabilities in an unsupervised manner. We then apply the method to real stellar streams in \autoref{sec:results_gd1} and \autoref{sec:results_pal5}.

@@

    We now apply our density modeling approach to the stellar stream GD-1. We discuss the data selection in \autoref{sub:results_gd1_data} and results in \autoref{ssub:trained_gd1}.

    \subsection{Data Selection} \label{sub:results_gd1_data}

@@

    \subsection{Model Specification}\label{sec:GD_1_ModelSpecification}

@@

    \subsection{Trained Density Model}\label{ssub:trained_gd1}

@@

\section{Palomar 5} \label{sec:results_pal5}

@@

% section pal5 (end)
